Robustly decorrelating errors with mixed quantum gates

### 2019-09-16 decorrelation/mixed_unitaries.tex
Minor changes to tex file

CPTP maps possess a number of useful representations, including Kraus operators\cite{1983}, Choi matrices\cite{Choi1975}, and Jamiolkowsi states\cite{yczkowski2004}. But for the purposes of this article, the \emph{process matrix} representation will be particularly convenient\cite{OBrien2004}, and we shall denote the process matrix associated with a given CPTP map $\target$ with its corresponding calligraphic character, $\tarmat$. For a $d$-dimensional quantum state, the process matrix is a $d^2\times d^2$ matrix. A key feature of process matrices is that they are composable and act through the usual matrix multiplication on the vectorized quantum state:
\begin{align}
	\vectorize{\actual(\rho)}
		&= \actmat\cdot\vectorize{\rho} \\
		&= \errmat\cdot\tarmat\cdot\vectorize{\rho}
\end{align}
The $\mathsf{vec}$ operation is typically performed using a basis of matrix units, for which $\vectorize{\rho}$ would be the column vector obtained by stacking the columns of $\rho$. In this work, however, we shall use a basis of Pauli matrices, defining $P = \left\{I, \sigma_x, \sigma_y, \sigma_z\right\}^{\otimes n}$ as the collection of all $4^n$ $n$-qubit Pauli operators (including the identity).  In this basis,
\begin{equation}
  \vectorize{\rho}_i = \expect{P_i} = \tr\!\left(P_i\; \rho \right),
\end{equation}
and
\begin{equation}
	(\actmat)_{ij} = \tr\!\left(P_i \;\actual\!\left(P_j\right) \right).
\end{equation}
Process matrices written in this Pauli basis are often referred to as \emph{Pauli transfer matrices} \cite{Chow2012}.  Error maps represented in this basis take the particularly nice form,
\begin{equation}\label{eq:process_matrix}
\errmat =
	\left(\begin{array}{c|cccc}
		1 &  & \vec{0}^T & \\ 
		\hline & &  &  \\
		\vec{m} &  & R &  \\
		 &  &  & 
	\end{array} 	
	\right)
\end{equation}
The top row of all trace-preserving (TP) error maps is fixed to $\{1,0,0,0,\cdots\}$ and the remainder of the first column, $\vec{m}$, describes any deviations from unitality, as might arise from amplitude damping \cite{preskill1997lecture}. If the error map is unitary, then the error is called \textit{coherent} and the unital submatrix $R$ is a rotation matrix.  Importantly, if $R$ is diagonal, then the error is Pauli stochastic, with each diagonal entry corresponding to the probability that its associated Pauli error occurs in each application of the gate. \kcy{We might need more here defining coherent noise. There is a lot of literature on this, most of it inconsistent. Try Joel's Estimating the Coherence of Noise (terrible title) and Gottesman's paper on Maximally Sensitive States.}

@@

\subsubsection{Robustness to drift and model uncertainty} % (fold)
\label{sub:adding_robustness}
\noindent 
While mixed quantum gates offer significant performance improvements, their construction assumes a good knowledge of the errors experienced by the constituent gates. But often the gates are not perfectly well characterized. This could be due to simple uncertainty about the parameters of the model used to generate the control solutions, or the system could experience some degree of drift \kcy{cite Tim's drift manuscript}. In either case, the performance of the MQG will be negatively impacted. We can describe the effect of this uncertainty by first writing the gate explicitly as a function of a vector of model parameters, $\vec\lambda$, and their nominal values, $\vec\lambda_0$. These model parameters could include magnetic field strengths, laser or microwave intensities, coupling strengths, etc. If the parameters are close to their nominal values, then we can write a gate's error generator as a Taylor expansion in the deviation vector, $\vec\delta = \vec\lambda-\vec\lambda_0$:
\begin{equation}
	\label{eq:sensitivity_expansion}
	\genmat_i(\vec \delta) = \genmat_i(\0) + \sum_k \delta_k \,\genmat_{i,k}(\0) + \order{\delta^2}.
\end{equation} 
In the above expression we have used the comma derivative, 
\begin{equation}
	\genmat_{i,k}(\vec \delta_0) = \frac{\partial}{\partial \delta_k} \genmat_i(\vec\delta) \vert_{\vec \delta=\delta_0}.
\end{equation}

Combining Eq.~\eqref{eq:sensitivity_expansion} with Eq.~\eqref{eq:generator} and taking both Taylor expansions to first order, we have:
\begin{align}
\notag
	\errmat_{\rm eff} =&
		 \identmat_d + \sum_i \weight_i \left(\genmat_i(0) + \sum_k \delta_k \,\genmat_{i,k}(\0) \right) \\
		 &+ \order{\genmat_i^2 + \genmat_i \delta + \delta^2}.
\end{align}
We would like to choose the weights $\vec\weight$ so that the effective error generator vanishes to first order. We can cast this problem as a convex optimization problem, much as we did in Sec.~\ref{sec:introduction} for synthesizing generator-exact MQGs. Following Eq.~\eqref{eq:minimization}, we have,
\begin{equation}\label{eq:minimization}
  \begin{split}
    &\underset{\weight_i\geq0, |\weight|_1=1}{\textbf{minimize}: } ||\mathbf{L}\cdot\vec\weight||_2 + \sum_k||\mathbf{L}_{,k}\cdot\vec\weight||_2\\
  \end{split}
\end{equation}
Where $\mathbf{L}_{,k}=\left( \mathsf{col}(\genmat_{1,k}(\0)) \; \mathsf{col}(\genmat_{2,k}(\0)) \; \cdots \;  \right)$ is the $d^2\times m$-dimensional matrix of column-stacked derivatives of the error generators with respect to the deviation $\delta_k$.

@@

\subsubsection{Sparsity Constraints}
\label{sec:sparsity}
As a practical consideration, we would also like to regularize our objective function to promote sparse weightings. Control electronics often have a limited amount of waveform memory, and thus it is important to be able to construct MQGs as few constituent gates as possible. In many machine learning contexts, lasso regularization \cite{tibshirani1996regression} can be used to enforce sparsity in solutions. Rather than penalizing the $\ell_0$ norm of the solution, which is not convex, lasso regularization penalizes the $\ell_1$, which is the convex relaxation of the $\ell_0$ norm. However this is insufficient for our purposes, as we require $\weight$ to be a valid probability distribution, which constrains the $\ell_1$ norm to be equal to $1$. Conveniently, the problem of enforcing sparsity in such situations has been considered in \cite{NIPS2012_4504} and can be expressed via another convex program that extends (\ref{eq:robust_minimization}):

@@

\subsection{Experimental implementation} % (fold)
\label{sub:experimental}
\noindent We implemented our methods using the 19Q-Acorn superconducting transmon processor at Rigetti Computing. Comprehensive device characterization for this processor can be found in \cite{1712.05771}. While this device consisted of 20 qubits with fixed couplings, we used only qubit \#8 for our demonstration. Controls were derived from a single, 10-sample, 50ns Gaussian $X_{\pi/2}$pulse calibrated by iteratively amplifying and correcting over- and under-rotation errors via repeated application. Four intentionally miscalibrated Gaussian pulses, labeled $\mathsf{Pulse1}$ through $\mathsf{Pulse4}$, were derived from this waveform by scaling the amplitude of the calibrated $X_{\pi/2}$ pulse by $106.4\%$,  $103.9\%$, $93.7\%$ and $91.2\%$. The gates therefore experience unitary over/under rotations, which are coherent errors. These gates are then used as in Sec.~\ref{sub:off_diagonals} to construct a Pauli-exact MQG, which samples from the miscalibrated waveforms with probabilities, \kcy{What are they?}, respectively.

>>>>

CPTP maps possess a number of useful representations, including Kraus operators\cite{1983}, Choi matrices\cite{Choi1975}, and Jamiolkowsi states\cite{yczkowski2004}. But for the purposes of this article, the \emph{process matrix} representation will be particularly convenient\cite{OBrien2004}, and we shall denote the process matrix associated with a given CPTP map $\target$ with its corresponding calligraphic character, $\tarmat$. For a $d$-dimensional quantum state, the process matrix is a $d^2\times d^2$ matrix. A key feature of process matrices is that they are composable and act through the usual matrix multiplication on the vectorized quantum state:
\begin{align}
	\vectorize{\actual(\rho)}
		&= \actmat\cdot\vectorize{\rho} \\
		&= \errmat\cdot\tarmat\cdot\vectorize{\rho}
\end{align}
The $\mathsf{vec}$ operation is typically performed using a basis of matrix units, for which $\vectorize{\rho}$ would be the column vector obtained by stacking the columns of $\rho$. In this work, however, we shall use a basis of Pauli matrices, defining $P = \left\{I, \sigma_x, \sigma_y, \sigma_z\right\}^{\otimes n}$ as the collection of all $4^n$ $n$-qubit Pauli operators (including the identity).  In this basis,
\begin{equation}
  \vectorize{\rho}_i = \expect{P_i} = \tr\!\left(P_i\; \rho \right),
\end{equation}
and
\begin{equation}
	(\actmat)_{ij} = \frac{1}{d}\tr\!\left(P_i \;\actual\!\left(P_j\right) \right).
\end{equation}
Process matrices written in this Pauli basis are often referred to as \emph{Pauli transfer matrices} \cite{Chow2012}.  Error maps represented in this basis take the particularly nice form,
\begin{equation}\label{eq:process_matrix}
\errmat =
	\left(\begin{array}{c|cccc}
		1 &  & \vec{0}^T & \\ 
		\hline & &  &  \\
		\vec{m} &  & R &  \\
		 &  &  & 
	\end{array} 	
	\right)
\end{equation}
The top row of all trace-preserving (TP) error maps is fixed to $\{1,0,0,0,\cdots\}$ and the remainder of the first column, $\vec{m}$, describes any deviations from unitality, as might arise from amplitude damping \cite{preskill1997lecture}. If the error map is unitary, then the error is called \textit{coherent} and the unital submatrix $R$ is a rotation matrix.  Importantly, if $R$ is diagonal, then the error is Pauli stochastic, with each diagonal entry corresponding to the probability that its associated Pauli error occurs in each application of the gate. \kcy{We might need more here defining coherent noise. There is a lot of literature on this, most of it inconsistent. Try Joel's Estimating the Coherence of Noise (terrible title) and Gottesman's paper on Maximally Sensitive States.}

@@

\subsubsection{Robustness to drift and model uncertainty} % (fold)
\label{sub:adding_robustness}
\noindent 
Mixed quantum gates can offer significant performance improvements over bare unitary gates, but their construction requires a good knowledge of the errors experienced by the constituent gates. Often, however, the gates are not perfectly well characterized. This could be due to simple uncertainty about the parameters of the model used to generate the control solutions, or the system could experience some degree of drift \kcy{cite Tim's drift manuscript}. In either case, the performance of the MQG will be negatively impacted. We can describe the effect of this uncertainty by first writing the gate explicitly as a function of a vector of model parameters, $\vec\lambda$, and their nominal values, $\vec\lambda_0$. These model parameters could include magnetic field strengths, laser or microwave intensities, coupling strengths, etc. If the parameters are close to their nominal values, then we can write a gate's error generator as a Taylor expansion in the deviation vector, $\vec\delta = \vec\lambda-\vec\lambda_0$:
\begin{equation}
	\label{eq:sensitivity_expansion}
	\genmat_i(\vec \delta) = \genmat_i(\0) + \sum_k \delta_k \,\genmat_{i,k}(\0) + \order{\delta^2}.
\end{equation} 
In the above expression we have used the comma derivative, 
\begin{equation}
	\genmat_{i,k}(\vec \delta_0) = \frac{\partial}{\partial \delta_k} \genmat_i(\vec\delta) \vert_{\vec \delta=\delta_0}.
\end{equation}

Combining Eq.~\eqref{eq:sensitivity_expansion} with Eq.~\eqref{eq:generator} and taking both Taylor expansions to first order, we have:
\begin{align}
\notag
	\errmat_{\rm eff} =
		 \identmat_d &+ \sum_i \weight_i \left(\genmat_i(\0) + \sum_k \delta_k \,\genmat_{i,k}(\0) \right) \\
		 &+ \order{\genmat_i^2 + \genmat_i \delta + \delta^2}.
\end{align}
We would like to choose the weights $\vec\weight$ so that the effective error generator vanishes to first order. We can cast this problem as a convex optimization problem, much as we did in Sec.~\ref{sec:introduction} for synthesizing generator-exact MQGs. Following Eq.~\eqref{eq:minimization}, we have,
\begin{equation}\label{eq:minimization}
  \begin{split}
    &\underset{\weight_i\geq0, |\weight|_1=1}{\textbf{minimize}: } ||\mathbf{L}\cdot\vec\weight||_2 + \sum_k||\mathbf{L}_{,k}\cdot\vec\weight||_2\\
  \end{split}
\end{equation}
Where $\mathbf{L}_{,k}=\left( \mathsf{col}(\genmat_{1,k}(\0)) \; \mathsf{col}(\genmat_{2,k}(\0)) \; \cdots \;  \right)$ is the $d^2\times m$-dimensional matrix of column-stacked derivatives of the error generators with respect to the deviation $\delta_k$.

@@

\subsubsection{Sparsity Constraints}
\label{sec:sparsity}
\noindent
As a practical consideration, we would also like to regularize our objective function to promote sparse weightings. Control electronics often have a limited amount of waveform memory, and thus it is important to be able to construct MQGs as few constituent gates as possible. In many machine learning contexts, lasso regularization \cite{tibshirani1996regression} can be used to enforce sparsity in solutions. Rather than penalizing the $\ell_0$ norm of the solution, which is not convex, lasso regularization penalizes the $\ell_1$, which is the convex relaxation of the $\ell_0$ norm. However this is insufficient for our purposes, as we require $\weight$ to be a valid probability distribution, which constrains the $\ell_1$ norm to be equal to $1$. Conveniently, the problem of enforcing sparsity in such situations has been considered in \cite{NIPS2012_4504} and can be expressed via another convex program that extends (\ref{eq:robust_minimization}):

@@

\subsection{Experimental implementation } % (fold)
\label{sub:experimental}
\noindent We implemented our methods using the 19Q-Acorn superconducting transmon processor at Rigetti Computing. Comprehensive device characterization for this processor can be found in \cite{1712.05771}. While this device consisted of 20 qubits with fixed couplings, we used only qubit \#8 for our demonstration. Controls were derived from a single, 10-sample, 50ns Gaussian $X_{\pi/2}$pulse calibrated by iteratively amplifying and correcting over- and under-rotation errors via repeated application. Four intentionally miscalibrated Gaussian pulses, labeled $\mathsf{Pulse1}$ through $\mathsf{Pulse4}$, were derived from this waveform by scaling the amplitude of the calibrated $X_{\pi/2}$ pulse by $106.4\%$,  $103.9\%$, $93.7\%$ and $91.2\%$. The gates therefore experience unitary over/under rotations, which are coherent errors. These gates are then used as in Sec.~\ref{sub:off_diagonals} to construct a Pauli-exact MQG, which samples from the miscalibrated waveforms with probabilities, \kcy{What are they?}, respectively.
